---
title: "Population Surface Construction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Population Surface Construction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

# Introduction

When `spseg()` is called with `surface = "grid"` or `surface = "pycno"`,
polygon counts are first moved to a regular square grid. This vignette walks
through that step with small examples. The goal is to make clear what happens
when every polygon receives grid cells directly, and what happens when one or
more populated polygons are too small to receive a grid-cell centroid.

```{r setup}
library(seg)
library(sf)
library(dplyr)
library(ggplot2)

composition_scale <- function(name = "A share") {
  scale_fill_gradient2(
    low = "#f4a6a6",
    mid = "#eee6ff",
    high = "#a8c7ff",
    midpoint = 0.5,
    limits = c(0, 1),
    name = name
  )
}

make_rect <- function(xmin, ymin, xmax, ymax) {
  st_polygon(list(rbind(
    c(xmin, ymin),
    c(xmax, ymin),
    c(xmax, ymax),
    c(xmin, ymax),
    c(xmin, ymin)
  )))
}

make_irregular_case <- function(edge_width) {
  st_sf(
    A = c(12, 0, 60, 15),
    B = c(0, 8, 30, 75),
    geometry = st_sfc(
      make_rect(0, 0, edge_width, 1),
      make_rect(0, 5, edge_width, 6),
      st_polygon(list(rbind(
        c(edge_width, 0),
        c(3, 0),
        c(4, 6),
        c(edge_width, 6),
        c(edge_width, 5),
        c(0, 5),
        c(0, 1),
        c(edge_width, 1),
        c(edge_width, 0)
      ))),
      st_polygon(list(rbind(
        c(3, 0), c(6, 0), c(6, 6), c(4, 6), c(3, 0)
      )))
    )
  ) |>
    source_to_sf()
}

surface_from_polygons <- function(x, nrow, ncol, surface = "grid") {
  checked <- suppressMessages(seg:::chksegdata(x[, c("A", "B")]))
  seg:::spseg_surface(
    x = x[, c("A", "B")],
    coords = checked$coords,
    data = checked$data,
    surface = surface,
    args = list(
      nrow = nrow,
      ncol = ncol,
      surface_geometry = "polygons",
      converge = 1
    ),
    verbose = FALSE
  )
}

surface_to_sf <- function(surface) {
  st_sf(
    cell = seq_len(nrow(surface$data)),
    source_id = surface$id,
    as.data.frame(surface$data),
    geometry = surface$geometry
  ) |>
    mutate(
      fallback_cell = cell %in% surface$fallback$cells,
      total = A + B,
      A_share = A / total
    )
}

source_to_sf <- function(x) {
  x |>
    mutate(
      source_id = row_number(),
      total = A + B,
      A_share = A / total
    )
}

plot_sources <- function(x, title) {
  centroids <- st_point_on_surface(st_geometry(x))
  labels <- cbind(st_drop_geometry(x), st_coordinates(centroids))

  ggplot(x) +
    geom_sf(aes(fill = A_share), color = "white", linewidth = 0.8) +
    geom_text(
      data = labels,
      aes(X, Y, label = paste0("id ", source_id, "\nA=", A, "\nB=", B)),
      size = 3
    ) +
    coord_sf(expand = FALSE) +
    composition_scale() +
    labs(title = title) +
    theme_minimal()
}

plot_surface <- function(x, title, labels = c("all", "fallback")) {
  labels <- match.arg(labels)
  centroids <- st_point_on_surface(st_geometry(x))
  label_data <- cbind(st_drop_geometry(x), st_coordinates(centroids)) |>
    mutate(
      label = if (labels == "all") {
        paste0("cell ", cell, "\nsource ", source_id,
               "\nA=", round(A, 1), "\nB=", round(B, 1))
      } else if_else(
        fallback_cell,
        paste0("cell ", cell, "\nA=", round(A, 1), "\nB=", round(B, 1)),
        as.character(cell)
      )
    )

  ggplot(x) +
    geom_sf(aes(fill = A_share), color = "white", linewidth = 0.8) +
    geom_sf(
      data = filter(x, fallback_cell),
      fill = NA,
      color = "#333333",
      linewidth = 1.4
    ) +
    geom_text(
      data = label_data,
      aes(X, Y, label = label),
      size = if (labels == "all") 2.8 else 2.5
    ) +
    coord_sf(expand = FALSE) +
    composition_scale("Grid A share") +
    labs(title = title) +
    theme_minimal()
}

plot_surface_change <- function(x, title) {
  limit <- max(abs(x$A_share_change))

  ggplot(x) +
    geom_sf(aes(fill = A_share_change), color = "white", linewidth = 0.8) +
    geom_sf_text(
      aes(label = paste0(cell, "\n", sprintf("%+.2f", A_share_change))),
      size = 2.4
    ) +
    coord_sf(expand = FALSE) +
    scale_fill_gradient2(
      low = "#f4a6a6",
      mid = "white",
      high = "#a8c7ff",
      midpoint = 0,
      limits = c(-limit, limit),
      name = "Change in\nA share"
    ) +
    labs(title = title) +
    theme_minimal()
}

count_check <- function(source, surface) {
  tibble(
    group = c("A", "B"),
    source_total = colSums(st_drop_geometry(source[, c("A", "B")])),
    surface_total = colSums(surface$data),
    equal = abs(source_total - surface_total) < 1e-8
  )
}
```

# Case 1: No Fallback Required

Start with a six-by-six grid over four irregular neighborhoods. Two larger
neighborhoods are separated by a slanted boundary. Two smaller neighborhoods
sit at opposite ends of the left edge. Their width is `0.75`, so each contains
one of the grid centroids at `x = 0.5`. Every populated neighborhood therefore
receives at least one cell during first-pass rasterization.

```{r no-merge-data}
no_fallback <- make_irregular_case(edge_width = 0.75)
no_fallback_grid <- surface_from_polygons(
  no_fallback,
  nrow = 6,
  ncol = 6
)
no_fallback_surface <- surface_to_sf(no_fallback_grid)
```

```{r no-merge-source-map}
plot_sources(no_fallback, "Irregular source polygons: no fallback required")
```

```{r no-merge-grid-map}
plot_surface(
  no_fallback_surface,
  "Grid surface: all neighborhoods assigned on the first pass",
  labels = "fallback"
)
```

The larger neighborhoods are represented by multiple cells, so their counts are
distributed evenly across those cells. Summing the cells by `source_id`
reproduces every source-neighborhood count, and the fallback metadata is empty.

```{r no-merge-check}
count_check(no_fallback, no_fallback_grid)

no_fallback_grid$fallback

no_fallback_surface |>
  st_drop_geometry() |>
  group_by(source_id) |>
  summarise(A = sum(A), B = sum(B), .groups = "drop") |>
  left_join(
    no_fallback |>
      st_drop_geometry() |>
      select(source_id, A_source = A, B_source = B),
    by = "source_id"
  )
```

# Pycnophylactic Smoothing Without Fallback

The gridded surface is piecewise constant before smoothing: all cells assigned
to a given source neighborhood have the same counts. Pycnophylactic smoothing
uses neighboring cells to soften these abrupt changes, then rescales cells
within each source zone so that every neighborhood's original group totals are
restored.

```{r no-fallback-pycno}
no_fallback_pycno <- surface_from_polygons(
  no_fallback,
  nrow = 6,
  ncol = 6,
  surface = "pycno"
)
no_fallback_pycno_surface <- surface_to_sf(no_fallback_pycno)

no_fallback_change <- no_fallback_pycno_surface |>
  rename(A_share_pycno = A_share) |>
  left_join(
    no_fallback_surface |>
      st_drop_geometry() |>
      select(cell, A_share_grid = A_share),
    by = "cell"
  ) |>
  mutate(A_share_change = A_share_pycno - A_share_grid)
```

```{r no-fallback-pycno-map}
plot_surface(
  no_fallback_pycno_surface,
  "Pycnophylactic surface: no fallback required",
  labels = "fallback"
)
```

```{r no-fallback-change-map}
plot_surface_change(
  no_fallback_change,
  "Where pycnophylactic smoothing changes group A share"
)
```

The largest changes occur around the slanted boundary and in the large
neighborhood near the two small corner neighborhoods. The corner neighborhoods
are each represented by only one grid cell. Because pycno must preserve their
zonal totals, those cells remain fixed. Their composition nevertheless affects
neighboring cells during the averaging step.

```{r no-fallback-pycno-check}
grid_zone_totals_no_fallback <- no_fallback_surface |>
  st_drop_geometry() |>
  group_by(source_id) |>
  summarise(across(c(A, B), sum), .groups = "drop")

pycno_zone_totals_no_fallback <- no_fallback_pycno_surface |>
  st_drop_geometry() |>
  group_by(source_id) |>
  summarise(across(c(A, B), sum), .groups = "drop")

left_join(
  grid_zone_totals_no_fallback,
  pycno_zone_totals_no_fallback,
  by = "source_id",
  suffix = c("_grid", "_pycno")
)

no_fallback_change |>
  st_drop_geometry() |>
  filter(source_id %in% c(1, 2) | cell %in% c(2, 7, 25, 32)) |>
  select(cell, source_id, A_share_grid, A_share_pycno, A_share_change)
```

# Case 2: Fallback Assignment Required

Keep the same six-by-six grid, group counts, and overall geometric scheme, but
narrow the two edge neighborhoods from `0.75` to `0.25`. They no longer contain
the grid centroids at `x = 0.5`, so the first-pass rasterization misses them.
The two larger neighborhoods retain their irregular shapes and slanted shared
boundary.

The fallback step finds a representative interior point for each missed polygon
and identifies the grid cell containing that point. The missing polygon's counts
are added directly to that cell. They are not spread over the other cells
assigned to the larger polygon. This retains useful information about where the
small polygon was located, while still preserving the overall group counts.

```{r merge-data}
fallback_case <- make_irregular_case(edge_width = 0.25)

fallback_grid <- surface_from_polygons(fallback_case, nrow = 6, ncol = 6)
fallback_surface <- surface_to_sf(fallback_grid)
```

```{r merge-source-map}
plot_sources(fallback_case, "Irregular source polygons")
```

```{r merge-grid-map}
plot_surface(
  fallback_surface,
  "Grid surface after fallback assignment",
  labels = "fallback"
)
```

The fallback metadata reports which source polygons were missed by the initial
centroid assignment and which fallback cells received their counts. The fallback
cells are outlined in dark gray in the grid map.

```{r merge-fallback}
tibble(
  missing_source_id = fallback_grid$fallback$ids,
  fallback_cell = fallback_grid$fallback$cells
)
```

The overall group counts are still preserved.

```{r merge-count-check}
count_check(fallback_case, fallback_grid)
```

The source-to-grid bookkeeping still changes. Source polygons 1 and 2 are not
represented as separate grid zones because their fallback cells already belong
to source zone 3. Their counts therefore contribute to the effective total for
zone 3. Unlike the merge-and-redistribute approach, however, the counts remain
concentrated in the two fallback cells. The table contrasts those cells with
the ordinary cells assigned to zone 3.

```{r merge-zone-check}
source_totals <- fallback_case |>
  st_drop_geometry() |>
  select(source_id, A, B)

surface_totals <- fallback_surface |>
  st_drop_geometry() |>
  group_by(source_id) |>
  summarise(A = sum(A), B = sum(B), .groups = "drop")

source_totals
surface_totals

fallback_surface |>
  st_drop_geometry() |>
  filter(source_id == 3) |>
  group_by(fallback_cell) |>
  summarise(
    cells = paste(cell, collapse = ", "),
    A_min = min(A),
    A_max = max(A),
    B_min = min(B),
    B_max = max(B),
    .groups = "drop"
  )
```

# Pycnophylactic Smoothing Uses the Aggregated Grid

For `surface = "pycno"`, the same grid construction happens first. The
pycnophylactic smoother then operates on this grid. It no longer needs the
original polygon geometries, but it does use the effective grid-zone labels to
rebalance counts after each smoothing iteration. The initial fallback
concentrations provide a spatially faithful starting surface, while the
effective zone totals remain fixed. Iterative smoothing can diffuse those
cell-level concentrations and, with sufficient iteration, may largely remove
their initial contrast.

```{r pycno-merge}
fallback_pycno <- surface_from_polygons(
  fallback_case,
  nrow = 6,
  ncol = 6,
  surface = "pycno"
)
fallback_pycno_surface <- surface_to_sf(fallback_pycno)
```

```{r pycno-map}
plot_surface(
  fallback_pycno_surface,
  "Pycnophylactic surface after fallback assignment",
  labels = "fallback"
)
```

The overall counts and effective grid-zone totals remain unchanged after
smoothing.

```{r pycno-count-check}
count_check(fallback_case, fallback_pycno)

grid_zone_totals <- fallback_surface |>
  st_drop_geometry() |>
  group_by(source_id) |>
  summarise(across(c(A, B), sum), .groups = "drop")

pycno_zone_totals <- fallback_pycno_surface |>
  st_drop_geometry() |>
  group_by(source_id) |>
  summarise(across(c(A, B), sum), .groups = "drop")

left_join(
  grid_zone_totals,
  pycno_zone_totals,
  by = "source_id",
  suffix = c("_grid", "_pycno")
)

fallback_surface |>
  st_drop_geometry() |>
  filter(fallback_cell) |>
  select(cell, A_grid = A, B_grid = B) |>
  left_join(
    fallback_pycno_surface |>
      st_drop_geometry() |>
      filter(fallback_cell) |>
      select(cell, A_pycno = A, B_pycno = B),
    by = "cell"
  ) |>
  mutate(
    A_share_grid = A_grid / (A_grid + B_grid),
    A_share_pycno = A_pycno / (A_pycno + B_pycno)
  )
```

# Using the Surface in spseg()

In ordinary use, call `spseg()` directly. The same surface construction is used
internally before local environments and segregation indices are calculated.

```{r spseg-surface-call}
grid_result <- spseg(
  fallback_case[, c("A", "B")],
  surface = "grid",
  nrow = 6,
  ncol = 6,
  bands = c(1, 2),
  output = "full"
)

pycno_result <- spseg(
  fallback_case[, c("A", "B")],
  surface = "pycno",
  surface_audit = TRUE,
  nrow = 6,
  ncol = 6,
  bands = c(1, 2),
  output = "full",
  converge = 1
)

as.data.frame(grid_result, what = "indices")
as.data.frame(pycno_result, what = "indices")

pycno_result$surface_info$fallback
names(pycno_result$surface_info$audit)
head(pycno_result$surface_info$audit$cells)
pycno_result$surface_info$audit$fallback
```

The lightweight `surface_info$grid` and `surface_info$fallback` fields are
stored for gridded surfaces by default. Setting `surface_audit = TRUE` also
retains the source data and geometry, the mapping between full raster cells and
compact result rows, explicit fallback contributions, initial grid counts, and
smoothed counts. This audit information is retained even when
`output = "indices"`.
