---
title: "Empirical Example: Auckland"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Empirical Example: Auckland}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

This vignette uses empirical area-unit data for Auckland, New Zealand. The data
are included as external package files so the examples can be run after the
package is installed.

```{r setup}
devtools::load_all()
library(sf)
library(ggplot2)
```

# Load the Auckland Data

Files in `inst/extdata` are installed with the package. Use `system.file()` to
find them.

```{r data-paths}
auck_dir <- system.file("extdata", "auck", package = "seg")

if (auck_dir == "") {
  auck_dir <- file.path("..", "inst", "extdata", "auck")
}

auck_shp <- file.path(auck_dir, "auck.shp")
auck_pop <- file.path(auck_dir, "auck_pop.csv")

file.exists(auck_shp)
file.exists(auck_pop)
```

Read the shapefile with `sf`. The shapefile contains counts for Korean
residents and total population in 1991, 1996, 2001, and 2006.

```{r read-data}
auck <- sf::st_read(auck_shp, quiet = TRUE)
pop <- read.csv(auck_pop)

nrow(auck)
names(auck)
head(pop[, c("au_no", "au_name", "kr06", "pop06")])
```

For the examples below, define two mutually exclusive groups: Korean residents
and all other residents in 2006.

```{r groups}
auck$Korean <- auck$KR06
auck$Non_Korean <- auck$POP06 - auck$KR06
auck$korean_share <- auck$Korean / auck$POP06

summary(auck[, c("Korean", "Non_Korean", "korean_share")])
```

# Plot Raw Population Data

The raw map shows the share of residents identified as Korean in each area
unit.

```{r raw-map}
ggplot(auck) +
  geom_sf(aes(fill = korean_share), color = NA) +
  scale_fill_viridis_c(labels = function(x) paste0(round(100 * x, 1), "%")) +
  labs(fill = "Korean share") +
  theme_minimal()
```


# Calculate local environments and segregation indices.

`spseg()` produces both local environment data for specified bands as well as segregation indices based on those local environments. You can return the environments, the indices, or both. Here we specify `output='full'` to get both the indices and the local environments back. `search='kdtree'`, `neighbors='radius'`, and `measures='all'` are specified explicitly as well, but these are also the defaults. 

```{r localenv}
auck_seg <- spseg(
  auck[, c("Korean", "Non_Korean")],
  bands=c(1,1500,3000,4500,6000,9000),
  search="kdtree",
  neighbors="radius",
  measures="all",
  output="full"
)
auck_seg
```

# Plot Local Environment Data

`as.data.frame()` methods are available. Specifying `what='env'` and `shape='long'` returns the local environments in a tidy data.frame. 

```{r localenv-map}
auck_env<-as.data.frame(auck_seg,what="env",shape="long") 
auck_env$Total<-rowSums(auck_env[,c('Korean','Non_Korean')])
auck_env$Korean_Share <- auck_env$Korean/auck_env$Total+0.001

ggplot(auck_env) + 
  geom_sf(aes(geometry=geometry,fill = Korean_Share), color = NA) +
  scale_fill_viridis_c(labels = function(x) paste0(round(100 * x, 1), "%"), transform='log2') +
  facet_wrap(vars(band),ncol=2,
             labeller=as_labeller(
               function(x) paste('Bandwidth =',x,'m'))) +
  labs(fill = "Local Korean share") +
  theme_void()
```


# Bandwidth Sensitivity

Spatial segregation measures can change with the geographic scale used to
construct local environments. When we calculated the segregation measures, we already specified multiple bandwidths, we can extract the corresponding segregation measures using `as.data.frame()` with `what='indices'`. If we want the data in a tidy long format data frame, we can further specify `shape='long'`.

```{r bandwidth-sequence}
head(auck_ind<-as.data.frame(auck_seg,what="indices",shape="long"))
```

Then we can plot the resulting segregation curves:

```{r bandwidth-curve}
auck_ind |> 
ggplot(aes(x = band, y = value)) +
  geom_line() +
  geom_point(size = 1.5) +
  facet_wrap(~ measure, scales = "free_y") +
  labs(x = "Bandwidth (m)", y = "Measure value") +
  theme_minimal()
```

The exposure/isolation values are a bit more complited, but can be returned in long form using the `what='exposure'` option in `as.data.frame()`.

```{r exposure-curve}
head(auck_exp<-as.data.frame(auck_seg, what = "exposure") )

auck_exp |> 
ggplot(aes(x = band, y = exposure,color=from_group)) +
  geom_line() +
  geom_point(size = 1.5) +
  scale_y_continuous(labels=function(x) { format(100*x,digits=2) }) +
  facet_wrap(~ to_group, scales = "free_y",
             labeller=as_labeller(
               function(x) { paste("%",x) })) +
  labs(x = "Bandwidth (m)", y = "%",color='focal group') +
  theme_minimal()
```

```
