---
title: "Empirical Example: Auckland"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Empirical Example: Auckland}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

This vignette uses empirical area-unit data for Auckland, New Zealand. The data
are included as external package files so the examples can be run after the
package is installed.

```{r setup}
library(seg)
library(ggplot2)
```

# Load the Auckland Data

Files in `inst/extdata` are installed with the package. Use `system.file()` to
find them.

```{r data-paths}
auck_dir <- system.file("extdata", "auck", package = "seg")

if (auck_dir == "") {
  auck_dir <- file.path("..", "inst", "extdata", "auck")
}

auck_shp <- file.path(auck_dir, "auck.shp")
auck_pop <- file.path(auck_dir, "auck_pop.csv")

file.exists(auck_shp)
file.exists(auck_pop)
```

Read the shapefile with `sf`. The shapefile contains counts for Korean
residents and total population in 1991, 1996, 2001, and 2006.

```{r read-data}
auck <- sf::st_read(auck_shp, quiet = TRUE)
pop <- read.csv(auck_pop)

nrow(auck)
names(auck)
head(pop[, c("au_no", "au_name", "kr06", "pop06")])
```

For the examples below, define two mutually exclusive groups: Korean residents
and all other residents in 2006.

```{r groups}
auck$Korean <- auck$KR06
auck$Non_Korean <- auck$POP06 - auck$KR06
auck$korean_share <- auck$Korean / auck$POP06

summary(auck[, c("Korean", "Non_Korean", "korean_share")])
```

# Plot Raw Population Data

The raw map shows the share of residents identified as Korean in each area
unit.

```{r raw-map}
ggplot(auck) +
  geom_sf(aes(fill = korean_share), color = NA) +
  scale_fill_viridis_c(labels = function(x) paste0(round(100 * x, 1), "%")) +
  labs(fill = "Korean share") +
  theme_minimal()
```

The distribution is highly skewed, so a second map of the count is also useful.

```{r raw-count-map}
ggplot(auck) +
  geom_sf(aes(fill = Korean), color = NA) +
  scale_fill_viridis_c() +
  labs(fill = "Korean residents") +
  theme_minimal()
```

# Generate Local Environments

`localenv()` computes the local population composition around each area unit.
For projected coordinate systems measured in metres, the default exponential
decay parameter can be too steep. Here, `power = 0.0002` gives a smoother local
environment over neighbourhood-scale distances.

```{r localenv}
auck_env <- localenv(
  auck[, c("Korean", "Non_Korean")],
  power = 0.0002
)

auck_env
head(auck_env@env)
```

Add the local environment share back to the `sf` object for mapping.

```{r localenv-share}
auck$local_korean_share <- auck_env@env[, "Korean"] / rowSums(auck_env@env)

summary(auck$local_korean_share)
```

# Plot Local Environment Data

The local environment map smooths the raw area-unit shares by considering
nearby population composition.

```{r localenv-map}
ggplot(auck) +
  geom_sf(aes(fill = local_korean_share), color = NA) +
  scale_fill_viridis_c(labels = function(x) paste0(round(100 * x, 1), "%")) +
  labs(fill = "Local Korean share") +
  theme_minimal()
```

The two maps can also be compared directly.

```{r raw-vs-local}
plot_data <- rbind(
  data.frame(
    geometry = sf::st_geometry(auck),
    share = auck$korean_share,
    type = "Raw area-unit share"
  ),
  data.frame(
    geometry = sf::st_geometry(auck),
    share = auck$local_korean_share,
    type = "Local environment share"
  )
)
plot_data <- sf::st_as_sf(plot_data)

ggplot(plot_data) +
  geom_sf(aes(fill = share), color = NA) +
  facet_wrap(~ type) +
  scale_fill_viridis_c(labels = function(x) paste0(round(100 * x, 1), "%")) +
  labs(fill = "Korean share") +
  theme_minimal()
```

# Segregation Measures

The traditional index of dissimilarity is calculated from the raw population
counts.

```{r dissimilarity}
group_data <- sf::st_drop_geometry(auck[, c("Korean", "Non_Korean")])

dissim(data = group_data)
```

White's spatial proximity index can use the same `sf` object.

```{r proximity}
isp(auck[, c("Korean", "Non_Korean")])
```

Reardon and O'Sullivan's spatial measures can be calculated from the local
environment object.

```{r spatial-measures}
spatial_measures <- spatseg(auck_env)

as.list(spatial_measures)
```

For a compact comparison:

```{r measure-summary}
spatial_list <- as.list(spatial_measures)

data.frame(
  measure = c(
    "Dissimilarity",
    "Spatial proximity",
    "Spatial dissimilarity",
    "Relative diversity",
    "Information theory"
  ),
  value = c(
    dissim(data = group_data)$d,
    isp(auck[, c("Korean", "Non_Korean")]),
    spatial_list$d,
    spatial_list$r,
    spatial_list$h
  )
)
```

The exposure/isolation matrix is stored separately.

```{r exposure}
spatial_list$p
```
