---
title: "Measuring Spatial Segregation with seg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Measuring Spatial Segregation with seg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

The `seg` package implements segregation measures from the geographical and
sociological literature. These include Duncan and Duncan's index of
dissimilarity, White's index of spatial proximity, and Reardon and O'Sullivan's
spatial segregation measures.

This vignette demonstrates the current package interface using the bundled
`segdata` example data.

```{r setup}
library(seg)
```

# Example Data

The `segdata` data set contains eight hypothetical two-group population
patterns on a 10 by 10 grid. Each pattern uses two adjacent columns. For
example, columns `A1` and `A2` describe the first pattern.

```{r data}
data(segdata)

head(segdata[, 1:4])
```

For examples that require coordinates, create the centroids of a 10 by 10 grid.

```{r coordinates}
coords <- as.matrix(expand.grid(x = seq(0.5, 9.5), y = seq(0.5, 9.5)))
pattern_a <- segdata[, 1:2]

head(coords)
head(pattern_a)
```

# Index of Dissimilarity

Duncan and Duncan's index of dissimilarity, D, is computed with `dissim()`.
The function returns a list because spatially adjusted variants may also be
available.

```{r dissimilarity}
d_a <- dissim(data = pattern_a)
d_a

d_a$d
```

The first hypothetical pattern is completely segregated, so its traditional D
value is 1.

```{r all-dissimilarity}
vapply(seq_len(8), function(i) {
  idx <- 2 * i
  dissim(data = segdata[, (idx - 1):idx])$d
}, numeric(1))
```

# Spatial Proximity

White's index of spatial proximity is computed with `isp()`. If a distance
matrix is not supplied, distances are computed from the coordinates.

```{r spatial-proximity}
isp(coords, data = pattern_a)
```

The optional `fun` argument controls how distances are converted to proximity.
By default, `isp()` uses a negative exponential function.

```{r spatial-proximity-custom}
isp(coords, data = pattern_a, fun = function(d) 1 / (1 + d))
```

# Local Environments

The Reardon and O'Sullivan measures are based on local environments. Use
`localenv()` to calculate the local population composition around each point.
The ordinary coordinate-matrix path uses compiled C code for this calculation.

```{r local-environment}
env_a <- localenv(coords, data = pattern_a, maxdist = 2)

env_a
head(env_a@env)
```

When `useExp = FALSE` and `power = 0`, all points receive equal weight. In that
case every local environment equals the overall group means.

```{r local-environment-equal-weight}
env_equal <- localenv(coords, data = pattern_a, useExp = FALSE, power = 0)

head(env_equal@env)
colMeans(pattern_a)
```

# Reardon and O'Sullivan Measures

Use `spseg()` when you already have a `SegLocal` object.

```{r spseg-localenv}
spatial_a <- spseg(env_a)

as.list(spatial_a)
```

Use `spseg()` as a wrapper when you want to start from coordinates and
population data. By default, the package uses the compiled C implementation for
the index calculations.

```{r spseg}
spatial_a_2 <- spseg(coords, data = pattern_a, maxdist = 2)

as.list(spatial_a_2)
```

You can request a subset of measures with `method`.

```{r selected-methods}
selected <- spseg(
  coords,
  data = pattern_a,
  method = c("information", "dissimilarity"),
  maxdist = 2
)

as.list(selected)
```

# Working with sf Objects

Several functions also accept `sf` objects. The next example builds a 10 by 10
polygon grid and attaches `segdata` to it.

```{r sf-grid}
geom <- sf::st_polygon(
  list(rbind(c(0, 0), c(10, 0), c(10, 10), c(0, 10), c(0, 0)))
) |>
  sf::st_make_grid(cellsize = 1)

grid_sf <- sf::st_sf(segdata, geometry = geom)

nrow(grid_sf)
sf::st_geometry_type(grid_sf)[1]
```

If `data` is omitted, `seg` uses the non-geometry columns attached to the `sf`
object.

```{r sf-input}
isp(grid_sf[, c("A1", "A2")])

env_sf <- localenv(grid_sf[, c("A1", "A2")], maxdist = 2)
head(env_sf@env)
```

# Smoothing

The `spseg()` wrapper can compute segregation measures after constructing a
smoothed population surface. The `"equal"` option redistributes polygon
population values over a regular grid.

```{r equal-smoothing}
spseg(
  grid_sf,
  data = pattern_a,
  smoothing = "equal",
  nrow = 10,
  ncol = 10,
  maxdist = 2
) |>
  as.list()
```

The `"kernel"` option uses `splancs::kernel2d()` internally.

```{r kernel-smoothing}
window <- matrix(
  c(0, 0,
    0, 10,
    10, 10,
    10, 0),
  ncol = 2,
  byrow = TRUE
)

spseg(
  coords,
  data = pattern_a,
  smoothing = "kernel",
  nrow = 10,
  ncol = 10,
  window = window,
  sigma = 1,
  maxdist = 2
) |>
  as.list()
```

# Optional Spatial Adjustments

The `dissim()` function can calculate spatial adjustments to D when optional
packages are available. Morrill's adjustment uses `spdep`; Wong's shared
boundary adjustments use `terra`.

```{r optional-adjustments}
adjusted <- dissim(
  grid_sf,
  data = pattern_a,
  adjust = TRUE,
  use_queen = FALSE
)

adjusted
```

If the optional packages are not installed, unavailable adjusted measures are
returned as `NA`.
