---
title: "Empirical Example: Auckland"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Empirical Example: Auckland}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

This vignette uses empirical area-unit data for Auckland, New Zealand. The data
are included as external package files so the examples can be run after the
package is installed.

```{r setup}
library(seg)
library(sf)
library(ggplot2)
```

# Load the Auckland Data

Files in `inst/extdata` are installed with the package. Use `system.file()` to
find them.

```{r data-paths}
auck_dir <- system.file("extdata", "auck", package = "seg")

if (auck_dir == "") {
  auck_dir <- file.path("..", "inst", "extdata", "auck")
}

auck_shp <- file.path(auck_dir, "auck.shp")
auck_pop <- file.path(auck_dir, "auck_pop.csv")

file.exists(auck_shp)
file.exists(auck_pop)
```

Read the shapefile with `sf`. The shapefile contains counts for Korean
residents and total population in 1991, 1996, 2001, and 2006.

```{r read-data}
auck <- sf::st_read(auck_shp, quiet = TRUE)
pop <- read.csv(auck_pop)

nrow(auck)
names(auck)
head(pop[, c("au_no", "au_name", "kr06", "pop06")])
```

For the examples below, define two mutually exclusive groups: Korean residents
and all other residents in 2006.

```{r groups}
auck$Korean <- auck$KR06
auck$Non_Korean <- auck$POP06 - auck$KR06
auck$korean_share <- auck$Korean / auck$POP06

summary(auck[, c("Korean", "Non_Korean", "korean_share")])
```

# Plot Raw Population Data

The raw map shows the share of residents identified as Korean in each area
unit.

```{r raw-map}
ggplot(auck) +
  geom_sf(aes(fill = korean_share), color = NA) +
  scale_fill_viridis_c(labels = function(x) paste0(round(100 * x, 1), "%")) +
  labs(fill = "Korean share") +
  theme_minimal()
```


# Calculate local environments and segregation indices.

`spseg()` produces both local environment data for specified bands as well as segregation indices based on those local environments. You can return the environments, the indices, or both. Here we specify `output='full'` to get both the indices and the local environments back. `search='kdtree'`, `neighbors='radius'`, and `measures='all'` are specified explicitly as well, but these are also the defaults. 

```{r localenv}
auck_seg <- spseg(
  auck[, c("Korean", "Non_Korean")],
  bands=c(1,500,2500,5000),
  search="kdtree",
  neighbors="radius",
  measures="all",
  output="full"
)
auck_seg
```

# Plot Local Environment Data

`as.data.frame()` methods are available. Specifying `what='env'` and `shape='long'` returns the local environments in a tidy data.frame.

```{r localenv-map}
auck_env<-as.data.frame(auck_seg,what="env",shape="long") 
auck_env$Total<-colSums(auck_env[,c('Korean','Non_Korean')])
auck_env$Korean_Share <- auck_env$Korean/auck_env$Total

ggplot(auck_env) + 
  geom_sf(aes(fill = Korean_Share), color = NA) +
  scale_fill_viridis_c(labels = function(x) paste0(round(100 * x, 1), "%")) +
  facet_wrap(vars(band)) +
  labs(fill = "Local Korean share") +
  theme_minimal()
```

The two maps can also be compared directly.

```{r raw-vs-local}
plot_data <- rbind(
  data.frame(
    geometry = st_geometry(auck),
    share = auck$korean_share,
    type = "Raw area-unit share"
  ),
  data.frame(
    geometry = st_geometry(auck_env_sf),
    share = auck_env_sf$local_korean_share,
    type = "Local environment share"
  )
)
plot_data <- st_as_sf(plot_data)

ggplot(plot_data) +
  geom_sf(aes(fill = share), color = NA) +
  facet_wrap(~ type) +
  scale_fill_viridis_c(labels = function(x) paste0(round(100 * x, 1), "%")) +
  labs(fill = "Korean share") +
  theme_minimal()
```

# Bandwidth Sensitivity

Spatial segregation measures can change with the geographic scale used to
construct local environments. The following example calculates local
environments for bandwidths from 500 metres to 10 kilometres in 500 metre
increments.

```{r bandwidth-sequence}
bandwidths <- seq(500, 10000, by = 500)

bandwidth_result <- spseg(
  auck[, c("Korean", "Non_Korean")],
  bands = bandwidths,
  output = "full"
)
```

First, convert a subset of the stored local environments to an `sf` object for
faceted mapping.

```{r bandwidth-map-data}
map_bandwidths <- c(500, 2000, 5000, 10000)
bandwidth_maps <- st_as_sf(bandwidth_result, bands = map_bandwidths,
                           shape = "long")
bandwidth_counts <- st_drop_geometry(bandwidth_maps)[, c("Korean",
                                                         "Non_Korean")]
bandwidth_maps$local_korean_share <- bandwidth_maps$Korean /
  rowSums(bandwidth_counts)
bandwidth_maps$bandwidth <- paste0(bandwidth_maps$band / 1000, " km")

bandwidth_maps$bandwidth <- factor(
  bandwidth_maps$bandwidth,
  levels = paste0(map_bandwidths / 1000, " km")
)
```

```{r bandwidth-map, fig.height=6}
ggplot(bandwidth_maps) +
  geom_sf(aes(fill = local_korean_share), color = NA) +
  facet_wrap(~ bandwidth, ncol = 2) +
  scale_fill_viridis_c(labels = function(x) paste0(round(100 * x, 1), "%")) +
  labs(fill = "Local Korean share") +
  theme_minimal()
```

Next, calculate the segregation measures at every bandwidth.

```{r bandwidth-measures}
overall_measures <- as.data.frame(bandwidth_result, what = "indices")
overall_measures <- reshape(
  overall_measures[, c("band", "measure", "value")],
  idvar = "band",
  timevar = "measure",
  direction = "wide"
)
names(overall_measures) <- sub("^value\\.", "", names(overall_measures))

exposure_measures <- as.data.frame(bandwidth_result, what = "exposure")
korean_isolation <- exposure_measures[
  exposure_measures$from_group == "Korean" &
    exposure_measures$to_group == "Korean",
  c("band", "exposure")
]

bandwidth_measures <- merge(overall_measures, korean_isolation, by = "band")
names(bandwidth_measures)[names(bandwidth_measures) == "band"] <- "bandwidth_m"
names(bandwidth_measures)[names(bandwidth_measures) == "d"] <- "dissimilarity"
names(bandwidth_measures)[names(bandwidth_measures) == "r"] <- "diversity"
names(bandwidth_measures)[names(bandwidth_measures) == "h"] <- "information"
names(bandwidth_measures)[names(bandwidth_measures) == "exposure"] <-
  "korean_isolation"
bandwidth_measures$bandwidth_km <- bandwidth_measures$bandwidth_m / 1000

head(bandwidth_measures)
```

Plotting each measure against bandwidth gives a scale profile of segregation.

```{r bandwidth-curve}
curve_data <- reshape(
  bandwidth_measures,
  varying = c("dissimilarity", "diversity", "information", "korean_isolation"),
  v.names = "value",
  timevar = "measure",
  times = c(
    "Spatial dissimilarity",
    "Relative diversity",
    "Information theory",
    "Korean isolation"
  ),
  idvar = "bandwidth_km",
  direction = "long"
)

ggplot(curve_data, aes(x = bandwidth_km, y = value)) +
  geom_line() +
  geom_point(size = 1.5) +
  facet_wrap(~ measure, scales = "free_y") +
  labs(x = "Bandwidth (km)", y = "Measure value") +
  theme_minimal()
```

# Segregation Measures

The traditional index of dissimilarity is calculated from the raw population
counts.

```{r dissimilarity}
group_data <- sf::st_drop_geometry(auck[, c("Korean", "Non_Korean")])

dissim(data = group_data)
```

White's spatial proximity index can use the same `sf` object.

```{r proximity}
isp(auck[, c("Korean", "Non_Korean")])
```

Reardon and O'Sullivan's spatial measures can be calculated from the local
environment object.

```{r spatial-measures}
spatial_measures <- spseg(auck_env)

as.data.frame(spatial_measures, what = "indices")
```

For a compact comparison:

```{r measure-summary}
spatial_indices <- as.data.frame(spatial_measures, what = "indices")
spatial_values <- setNames(spatial_indices$value, spatial_indices$measure)

data.frame(
  measure = c(
    "Dissimilarity",
    "Spatial proximity",
    "Spatial dissimilarity",
    "Relative diversity",
    "Information theory"
  ),
  value = c(
    dissim(data = group_data)$d,
    isp(auck[, c("Korean", "Non_Korean")]),
    spatial_values["d"],
    spatial_values["r"],
    spatial_values["h"]
  )
)
```

The exposure/isolation matrix can be returned in long form.

```{r exposure}
as.data.frame(spatial_measures, what = "exposure")
```
