---
title: "Worked Example: A 3 by 3 Grid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Worked Example: A 3 by 3 Grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

# Introduction

This vignette walks through a very small example step by step. The goal is to
make the local environment calculations and the resulting segregation indices
easy to inspect.

The example uses a 3 by 3 grid of unit cells. Cell centroids are located at
half-unit coordinates. The upper-left three cells contain only group A, the
lower-right three cells contain only group B, and the remaining diagonal cells
contain equal counts of A and B.

```{r setup}
library(seg)
library(dplyr)
library(tidyr)
library(ggplot2)

composition_scale <- function(name = "A share") {
  scale_fill_gradient2(
    low = "#f4a6a6",
    mid = "#eee6ff",
    high = "#a8c7ff",
    midpoint = 0.5,
    limits = c(0, 1),
    name = name
  )
}
```

# Data

The grid is constructed so that `x` and `y` are the cell centroids. Because the
cells have unit dimensions, a bandwidth of 1 includes the focal cell and its
edge-adjacent neighbors. Diagonal neighbors are at distance `sqrt(2)` and are
therefore outside the bandwidth.

```{r grid-data}
grid <- tidyr::expand_grid(
  y = 2.5 - 0:2,
  x = 0.5 + 0:2
) |>
  mutate(
    id = row_number(),
    row = 3 - (y - 0.5),
    col = x + 0.5,
    A = case_when(
      row + col < 4 ~ 10,
      row + col == 4 ~ 5,
      TRUE ~ 0
    ),
    B = 10 - A
  ) |>
  select(id, row, col, x, y, A, B)

grid
```

```{r grid-map}
grid |>
  mutate(A_share = A / (A + B)) |>
  ggplot(aes(x = x, y = y, fill = A_share)) +
  geom_tile(color = "white", linewidth = 0.8, width = 1, height = 1) +
  geom_text(
    aes(x = x - 0.42, y = y + 0.38, label = id),
    inherit.aes = FALSE,
    size = 3,
    hjust = 0,
    vjust = 1
  ) +
  geom_text(aes(label = paste0("A=", A, "\nB=", B)), size = 3.5) +
  coord_equal() +
  composition_scale() +
  theme_minimal()
```

# Index Formulas

The examples below calculate two multigroup indices. With two groups, these
are also equal to their pairwise versions.

For each focal cell `i`, let `t_i` be the observed population count in the cell.
Let `p_iA` and `p_iB` be the local environment proportions for the two groups.
Let `P_A` and `P_B` be the regional proportions. In this example, every cell has
10 people and the region is exactly balanced, so `P_A = P_B = 0.5`.

The information theory index is

\[
H = 1 - \frac{\sum_i t_i E_i}{T E},
\]

where `E_i` is the entropy of the local environment and `E` is the regional
entropy.

The dissimilarity index is

\[
D = \frac{\sum_i t_i \sum_m |p_{im} - P_m|}{2 T I},
\]

where `I = sum_m P_m(1 - P_m)`.

The small helper below is intentionally written with ordinary `dplyr`
operations. It returns both the cell-level calculations and the final indices.

```{r index-helper}
xlogx <- function(x) {
  if_else(x > 0, x * log(x), 0)
}

calculate_indices <- function(local_env, label) {
  regional <- grid |>
    summarise(
      T_A = sum(A),
      T_B = sum(B),
      T = T_A + T_B
    ) |>
    mutate(
      P_A = T_A / T,
      P_B = T_B / T,
      E = -(xlogx(P_A) + xlogx(P_B)) / log(2),
      I = P_A * (1 - P_A) + P_B * (1 - P_B)
    )

  unit_calculations <- grid |>
    select(id, t_A = A, t_B = B) |>
    mutate(t_i = t_A + t_B) |>
    left_join(local_env, by = "id") |>
    mutate(
      local_total = A + B,
      p_A = A / local_total,
      p_B = B / local_total,
      E_i = -(xlogx(p_A) + xlogx(p_B)) / log(2),
      H_contribution = t_i * E_i,
      D_contribution = t_i * (
        abs(p_A - regional$P_A) + abs(p_B - regional$P_B)
      )
    )

  indices <- unit_calculations |>
    summarise(
      scenario = label,
      information = 1 - sum(H_contribution) / (regional$T * regional$E),
      dissimilarity = sum(D_contribution) / (2 * regional$T * regional$I)
    )

  list(
    regional = regional,
    units = unit_calculations,
    indices = indices
  )
}
```

# No Spatial Weighting

Without spatial weighting, each local environment is just the observed
composition of the focal cell. This is the same calculation that `spseg()`
performs when `bands = 0`.

```{r no-spatial-local-env}
raw_env <- grid |>
  select(id, A, B)

raw_env
```

```{r no-spatial-indices}
raw_calc <- calculate_indices(raw_env, "No spatial weighting")

raw_calc$regional
raw_calc$units |>
  select(id, t_i, A, B, p_A, p_B, E_i, H_contribution, D_contribution)
```

# Bandwidth 1 Without Distance Weights

Now use a radius bandwidth of 1. The first table lists every focal-neighbor
pair inside the bandwidth. The unweighted local environment gives each listed
neighbor, including the focal cell, equal weight.

```{r radius-neighbors}
neighbor_pairs <- grid |>
  select(focal_id = id, focal_x = x, focal_y = y) |>
  tidyr::crossing(
    grid |>
      select(neighbor_id = id, neighbor_x = x, neighbor_y = y, A, B)
  ) |>
  mutate(
    distance = sqrt(
      (focal_x - neighbor_x)^2 + (focal_y - neighbor_y)^2
    )
  ) |>
  filter(distance <= 1) |>
  arrange(focal_id, distance, neighbor_id)

neighbor_pairs
```

```{r unweighted-env}
unweighted_env <- neighbor_pairs |>
  mutate(weight = 1) |>
  group_by(id = focal_id) |>
  summarise(
    neighbors = paste(neighbor_id, collapse = ", "),
    A = sum(weight * A) / sum(weight),
    B = sum(weight * B) / sum(weight),
    .groups = "drop"
  )

unweighted_env
```

```{r unweighted-indices}
unweighted_calc <- calculate_indices(
  unweighted_env |> select(id, A, B),
  "Bandwidth 1, unweighted"
)

unweighted_calc$units |>
  select(id, t_i, A, B, p_A, p_B, E_i, H_contribution, D_contribution)
```

# Bandwidth 1 With Distance Weights

For the weighted example, use the exponential kernel. With bandwidth
normalization, the weight is `exp(-distance / bandwidth)`. Since the bandwidth
is 1, the focal cell has weight `exp(0) = 1` and each edge-adjacent neighbor
has weight `exp(-1)`.

```{r weighted-env}
weighted_env <- neighbor_pairs |>
  mutate(weight = exp(-distance / 1)) |>
  group_by(id = focal_id) |>
  summarise(
    neighbors = paste(neighbor_id, collapse = ", "),
    weight_sum = sum(weight),
    A = sum(weight * A) / weight_sum,
    B = sum(weight * B) / weight_sum,
    .groups = "drop"
  )

weighted_env
```

```{r weighted-indices}
weighted_calc <- calculate_indices(
  weighted_env |> select(id, A, B),
  "Bandwidth 1, exponential"
)

weighted_calc$units |>
  select(id, t_i, A, B, p_A, p_B, E_i, H_contribution, D_contribution)
```

# Compare with spseg()

The hand calculations above should match `spseg()` exactly. The calls below
calculate only the information theory and dissimilarity indices, using
multigroup scope to keep the output focused.

```{r spseg-comparison}
coords <- as.matrix(grid[, c("x", "y")])
counts <- as.matrix(grid[, c("A", "B")])

spseg_to_table <- function(result, scenario) {
  as.data.frame(result, what = "indices") |>
    select(measure, value) |>
    tidyr::pivot_wider(names_from = measure, values_from = value) |>
    transmute(
      scenario = scenario,
      information = h,
      dissimilarity = d
    )
}

spseg_raw <- spseg(
  coords,
  data = counts,
  bands = 0,
  weighting = "unweighted",
  measures = c("information", "dissimilarity"),
  scope = "multigroup"
)

spseg_unweighted <- spseg(
  coords,
  data = counts,
  bands = 1,
  weighting = "unweighted",
  measures = c("information", "dissimilarity"),
  scope = "multigroup",
  output = "full"
)

spseg_weighted <- spseg(
  coords,
  data = counts,
  bands = 1,
  weighting = "exponential",
  measures = c("information", "dissimilarity"),
  scope = "multigroup",
  output = "full"
)

manual_indices <- bind_rows(
  raw_calc$indices,
  unweighted_calc$indices,
  weighted_calc$indices
) |>
  mutate(source = "dplyr")

spseg_indices <- bind_rows(
  spseg_to_table(spseg_raw, "No spatial weighting"),
  spseg_to_table(spseg_unweighted, "Bandwidth 1, unweighted"),
  spseg_to_table(spseg_weighted, "Bandwidth 1, exponential")
) |>
  mutate(source = "spseg")
```

The next two figures compare the local environments produced by the manual
`dplyr` calculations with the local environments returned by `spseg()`. The
labels report the local environment counts, not the raw cell counts.

```{r local-env-plot-helper}
index_labels <- bind_rows(manual_indices, spseg_indices) |>
  mutate(
    facet_label = paste0(
      source,
      "\nH=", sprintf("%.3f", information),
      "\nD=", sprintf("%.3f", dissimilarity)
    )
  ) |>
  select(scenario, source, facet_label)

local_env_plot_data <- function(manual_env, spseg_result, scenario) {
  spseg_env <- as.data.frame(spseg_result$env[[1]])
  scenario_labels <- index_labels |>
    filter(.data$scenario == .env$scenario)

  bind_rows(
    grid |>
      select(id, x, y) |>
      left_join(manual_env |> select(id, A, B), by = "id") |>
      mutate(source = "dplyr"),
    grid |>
      select(id, x, y) |>
      mutate(
        A = spseg_env$A,
        B = spseg_env$B,
        source = "spseg"
      )
  ) |>
    left_join(
      scenario_labels,
      by = "source"
    ) |>
    mutate(
      facet_label = factor(
        facet_label,
        levels = scenario_labels |>
          arrange(match(source, c("dplyr", "spseg"))) |>
          pull(facet_label)
      ),
      A_share = A / (A + B),
      label = paste0(
        "A=", round(A, 2),
        "\nB=", round(B, 2)
      )
    )
}

plot_local_env_comparison <- function(plot_data) {
  ggplot(plot_data, aes(x = x, y = y, fill = A_share)) +
    geom_tile(color = "white", linewidth = 0.8, width = 1, height = 1) +
    geom_text(
      aes(x = x - 0.42, y = y + 0.38, label = id),
      inherit.aes = FALSE,
      size = 3,
      hjust = 0,
      vjust = 1
    ) +
    geom_text(aes(label = label), size = 3.2) +
    facet_wrap(~ facet_label) +
    coord_equal() +
    composition_scale("Local A share") +
    theme_minimal()
}
```

```{r unweighted-local-env-map}
unweighted_plot_data <- local_env_plot_data(
  unweighted_env,
  spseg_unweighted,
  "Bandwidth 1, unweighted"
)

plot_local_env_comparison(unweighted_plot_data)

tibble(
  local_environments_equal = isTRUE(all.equal(
    unweighted_plot_data |> filter(source == "dplyr") |> arrange(id) |>
      select(A, B),
    unweighted_plot_data |> filter(source == "spseg") |> arrange(id) |>
      select(A, B),
    tolerance = 1e-12
  ))
)
```

```{r weighted-local-env-map}
weighted_plot_data <- local_env_plot_data(
  weighted_env,
  spseg_weighted,
  "Bandwidth 1, exponential"
)

plot_local_env_comparison(weighted_plot_data)

tibble(
  local_environments_equal = isTRUE(all.equal(
    weighted_plot_data |> filter(source == "dplyr") |> arrange(id) |>
      select(A, B),
    weighted_plot_data |> filter(source == "spseg") |> arrange(id) |>
      select(A, B),
    tolerance = 1e-12
  ))
)
```

```{r equality-check}
comparison <- manual_indices |>
  select(-source) |>
  inner_join(
    spseg_indices |> select(-source),
    by = "scenario",
    suffix = c("_dplyr", "_spseg")
  ) |>
  mutate(
    information_equal = abs(information_dplyr - information_spseg) < 1e-12,
    dissimilarity_equal = abs(dissimilarity_dplyr - dissimilarity_spseg) < 1e-12
  )

comparison |>
  select(scenario, information_equal, dissimilarity_equal)
```
