---
title: "Empirical Example: Providence Race Segregation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Empirical Example: Providence Race Segregation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
population_path <- "/Users/bjarvis/Dropbox/Data Sources/US Census/Decennial/2010/RI_Block_Race/nhgis0026_ds172_2010_block.csv"
shape_path <- "/Users/bjarvis/Dropbox/Data Sources/Geography/United States/Census Shape Files/2010/tl2010_440_block_2010/RI_block_2010.shp"
have_ri_sources <- file.exists(population_path) && file.exists(shape_path)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = have_ri_sources
)
```

# Introduction

This vignette develops an empirical example using 2010 Census block data for
Providence, Rhode Island. It is intentionally written as a reproducible import
and preparation script: while the package is under development, the raw data are
read from local source files, joined to block geometries, recoded into mutually
exclusive race and Hispanic-origin categories, and then used to exercise the
multi-group and pairwise segregation workflows.

The population data come from IPUMS NHGIS table P5, Hispanic or Latino Origin
by Race, for 2010 Census blocks in Rhode Island. The block geometries are 2010
Census TIGER/Line block shapes for Rhode Island.

```{r source-check, eval = TRUE}
have_ri_sources
```

If the result above is `FALSE`, the local source files are not available and
the executable chunks below are skipped. The full import code is left in place
so the workflow can be rerun once the source files are available.

```{r setup}
library(seg)
library(sf)
library(ggplot2)
```

# Import Source Data

Read the NHGIS population extract and Rhode Island block geometries.

```{r read-source-data}

race <- read.csv(population_path, check.names = FALSE,
                 stringsAsFactors = FALSE)
race$Total <- race$H7Z001
race$White <- race$H7Z003
race$Black <- race$H7Z004
race$Asian <- race$H7Z006
race$Hispanic <- race$H7Z010
race$Other <- race$H7Z005 + race$H7Z007 + race$H7Z008 + race$H7Z009
race<-race[,c("GISJOIN","PLACEA","Total",'White','Black','Asian','Hispanic','Other')]

ri_blocks <- sf::st_read(shape_path, quiet = TRUE)

dim(race)
dim(ri_blocks)
names(race)
names(ri_blocks)
```

The NHGIS file includes a `PLACEA` code. Providence city is `59000`. The
shapefile and the NHGIS extract both include `GISJOIN`, which provides the
join key.

```{r filter-providence}
providence <- merge(ri_blocks, race[race$PLACEA==59000,], by = "GISJOIN")
providence <- providence[order(providence$GISJOIN), ]

nrow(providence)
sum(providence$Total)
unique(providence$COUNTY)
```

# Prepare Population Groups

The NHGIS P5 table separates Hispanic origin from race. For these examples,
construct mutually exclusive categories:

* `White`: not Hispanic or Latino, White alone
* `Black`: not Hispanic or Latino, Black or African American alone
* `Asian`: not Hispanic or Latino, Asian alone
* `Hispanic`: Hispanic or Latino, any race
* `Other`: remaining non-Hispanic categories

```{r prepare-groups}
group_cols <- c("White", "Black", "Asian", "Hispanic", "Other")
group_counts <- sf::st_drop_geometry(providence[, group_cols])

stopifnot(all(rowSums(group_counts) == providence$Total))
stopifnot(sum(providence$Total) == 178042)

colSums(group_counts)
```

Add a few shares for plotting and checking.

```{r group-shares}
providence$white_share <- ifelse(providence$Total > 0,
                                 providence$White / providence$Total, NA)
providence$black_share <- ifelse(providence$Total > 0,
                                 providence$Black / providence$Total, NA)
providence$asian_share <- ifelse(providence$Total > 0,
                                 providence$Asian / providence$Total, NA)
providence$hispanic_share <- ifelse(providence$Total > 0,
                                    providence$Hispanic / providence$Total, NA)

dominant_id <- max.col(group_counts, ties.method = "first")
providence$largest_group <- group_cols[dominant_id]
providence$largest_group <- ifelse(providence$Total<=0,NA,providence$largest_group)
providence$largest_group <- factor(providence$largest_group, levels = group_cols)


summary(providence[, c("Total", "white_share", "black_share",
                       "asian_share", "hispanic_share")])
```

# Plot Raw Data

The first map shows the largest group in each populated block.

```{r dominant-group-map}
ggplot(providence) +
  geom_sf(aes(fill = factor(largest_group)), color = NA) +
  scale_fill_discrete(na.value='gray70') +
  labs(fill = "Largest group") +
  theme_minimal()
```

The next map focuses on the Hispanic share of each block's population.

```{r hispanic-share-map}
ggplot(providence) +
  geom_sf(aes(fill = hispanic_share), color = NA) +
  scale_fill_viridis_c(labels = function(x) paste0(round(100 * x), "%"),na.value='gray70') +
  labs(fill = "Hispanic share") +
  theme_minimal()
```

# Local Environments

Providence block geometries are projected in meters, so bandwidths below are
interpreted as meters. A bandwidth of 1 meter is included as a reference case:
because block centroids are farther apart than this, it reproduces the original
block composition.

```{r local-environments}
env_bands <- c(1, 250, 500, 1000)
providence_env <- spseg(
  providence[, group_cols],
  bands = env_bands,
  weighting = "biweight",
  power = 3,
  normalize = TRUE,
  measures = "dissimilarity",
  comparison = "overall",
  output = "full"
)

providence_env
stopifnot(inherits(providence_env, "seg_result"))
stopifnot(length(providence_env$env) == length(env_bands))
```

Convert local environment counts to local Hispanic shares for mapping.

```{r local-share-data}
local_maps <- st_as_sf(providence_env, bands = env_bands, shape = "long")
local_maps$TotalRaw <- providence$Total
local_maps$Total <- rowSums(st_drop_geometry(local_maps)[, group_cols])
local_maps$hispanic_share <- local_maps$Hispanic / local_maps$Total
local_maps$hispanic_share <- ifelse(local_maps$TotalRaw <= 1e-12, NA,
                                    local_maps$hispanic_share)
local_maps$band <- factor(paste0(local_maps$band, " m"),
                          levels = paste0(env_bands, " m"))
summary(local_maps$hispanic_share)
```

```{r local-share-map}
ggplot(local_maps) +
  geom_sf(aes(fill = hispanic_share), color = NA) +
  facet_wrap(~ band,nrow=2) +
  scale_fill_viridis_c(labels = function(x) paste0(round(100 * x), "%"),na.value = 'gray70') +
  labs(fill = "Hispanic share") +
  theme_minimal()
```

# Multi-Band Segregation

Now compute multi-group segregation indices across several bandwidths. This
uses the unified multi-band engine and returns only the indices.

```{r multiband-indices}
bands <- c(250, 500, 1000, 2000, 4000, 6000, 8000, 10000)

providence_seg <- spseg(
  providence[, group_cols],
  bands = bands,
  weighting = "biweight",
  power = 3,
  normalize = TRUE,
  neighbors="radius",
  measures = c("dissimilarity", "information", "diversity"),
  comparison = "both",
  output = "indices"
)

providence_seg
stopifnot(inherits(providence_seg, "seg_result"))
stopifnot(identical(providence_seg$bands, bands))
stopifnot(is.null(providence_seg$env))
```

The overall indices summarize multi-group segregation at each bandwidth.

```{r overall-indices}
overall_indices <- as.data.frame(providence_seg, what = "indices")
overall_indices

stopifnot(nrow(overall_indices) == length(bands) * 3)
stopifnot(all(is.finite(overall_indices$value)))
```

```{r overall-index-plot}
ggplot(overall_indices, aes(x = band, y = value, color = measure)) +
  geom_line() +
  geom_point() +
  labs(x = "Bandwidth (m)", y = "Index value", color = "Measure") +
  theme_minimal()
```

# Pairwise Segregation

The same object also contains pairwise versions of the scalar indices. These
are returned separately from the overall multi-group indices.

```{r pairwise-indices}
pairwise_indices <- as.data.frame(providence_seg, what = "pairwise")
head(pairwise_indices)

expected_pairs <- choose(length(group_cols), 2)
stopifnot(nrow(pairwise_indices) == length(bands) * 3 * expected_pairs)
stopifnot(all(is.finite(pairwise_indices$value)))
```

```{r pairwise-dissimilarity-plot}
pairwise_d <- pairwise_indices[pairwise_indices$measure == "d", ]
pairwise_d$pair <- paste(pairwise_d$group_a, pairwise_d$group_b, sep = "-")

ggplot(pairwise_d[grepl("White",pairwise_d$pair),], 
       aes(x = band, y = value, color = pair)) +
  geom_line() +
  geom_point() +
  labs(x = "Bandwidth (m)", y = "Pairwise dissimilarity", color = "Pair") +
  theme_minimal()
```

# Source

Population data: IPUMS NHGIS, 2010 Census SF 1a, table P5, Hispanic or Latino
Origin by Race, Rhode Island blocks. Suggested citation from the NHGIS
codebook: Jonathan Schroeder, David Van Riper, Steven Manson, Katherine
Knowles, Tracy Kugler, Finn Roberts, and Steven Ruggles. IPUMS National
Historical Geographic Information System: Version 20.0 [dataset].
Minneapolis, MN: IPUMS. 2025. <http://doi.org/10.18128/D050.V20.0>.

Geometry: U.S. Census Bureau 2010 TIGER/Line block shapefile for Rhode Island.
