---
title: "Population Surface Construction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Population Surface Construction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

# Introduction

When `spseg()` is called with `surface = "grid"` or `surface = "pycno"`,
polygon counts are first moved to a regular square grid. This vignette walks
through that step with small examples. The goal is to make clear what happens
when every polygon receives grid cells directly, and what happens when one or
more populated polygons are too small to receive a grid-cell centroid.

```{r setup}
library(seg)
library(sf)
library(dplyr)
library(ggplot2)

composition_scale <- function(name = "A share") {
  scale_fill_gradient2(
    low = "#f4a6a6",
    mid = "#eee6ff",
    high = "#a8c7ff",
    midpoint = 0.5,
    limits = c(0, 1),
    name = name
  )
}

make_rect <- function(xmin, ymin, xmax, ymax) {
  st_polygon(list(rbind(
    c(xmin, ymin),
    c(xmax, ymin),
    c(xmax, ymax),
    c(xmin, ymax),
    c(xmin, ymin)
  )))
}

surface_from_polygons <- function(x, nrow, ncol, surface = "grid") {
  checked <- suppressMessages(seg:::chksegdata(x[, c("A", "B")]))
  seg:::spseg_surface(
    x = x[, c("A", "B")],
    coords = checked$coords,
    data = checked$data,
    surface = surface,
    args = list(
      nrow = nrow,
      ncol = ncol,
      surface_geometry = "polygons",
      converge = 1
    ),
    verbose = FALSE
  )
}

surface_to_sf <- function(surface) {
  st_sf(
    cell = seq_len(nrow(surface$data)),
    source_id = surface$id,
    as.data.frame(surface$data),
    geometry = surface$geometry
  ) |>
    mutate(total = A + B, A_share = A / total)
}

source_to_sf <- function(x) {
  x |>
    mutate(
      source_id = row_number(),
      total = A + B,
      A_share = A / total
    )
}

plot_sources <- function(x, title) {
  centroids <- st_point_on_surface(st_geometry(x))
  labels <- cbind(st_drop_geometry(x), st_coordinates(centroids))

  ggplot(x) +
    geom_sf(aes(fill = A_share), color = "white", linewidth = 0.8) +
    geom_text(
      data = labels,
      aes(X, Y, label = paste0("id ", source_id, "\nA=", A, "\nB=", B)),
      size = 3
    ) +
    coord_sf(expand = FALSE) +
    composition_scale() +
    labs(title = title) +
    theme_minimal()
}

plot_surface <- function(x, title) {
  centroids <- st_point_on_surface(st_geometry(x))
  labels <- cbind(st_drop_geometry(x), st_coordinates(centroids))

  ggplot(x) +
    geom_sf(aes(fill = A_share), color = "white", linewidth = 0.8) +
    geom_text(
      data = labels,
      aes(
        X, Y,
        label = paste0("cell ", cell, "\nsource ", source_id,
                       "\nA=", round(A, 1), "\nB=", round(B, 1))
      ),
      size = 2.8
    ) +
    coord_sf(expand = FALSE) +
    composition_scale("Grid A share") +
    labs(title = title) +
    theme_minimal()
}

count_check <- function(source, surface) {
  tibble(
    group = c("A", "B"),
    source_total = colSums(st_drop_geometry(source[, c("A", "B")])),
    surface_total = colSums(surface$data),
    equal = abs(source_total - surface_total) < 1e-8
  )
}
```

# Case 1: No Merge Required

Start with four square polygons and a two-by-two grid. Each polygon contains
one grid-cell centroid, so the first-pass rasterization assigns every polygon
to one cell. No fallback merge is needed.

```{r no-merge-data}
no_merge <- st_sf(
  A = c(10, 0, 5, 2),
  B = c(0, 10, 5, 8),
  geometry = st_sfc(
    make_rect(0, 0, 1, 1),
    make_rect(1, 0, 2, 1),
    make_rect(0, 1, 1, 2),
    make_rect(1, 1, 2, 2)
  )
) |>
  source_to_sf()

no_merge_grid <- surface_from_polygons(no_merge, nrow = 2, ncol = 2)
no_merge_surface <- surface_to_sf(no_merge_grid)
```

```{r no-merge-source-map}
plot_sources(no_merge, "Source polygons: no merge required")
```

```{r no-merge-grid-map}
plot_surface(no_merge_surface, "Grid surface: no merge required")
```

The cell counts are identical to the source counts because each source polygon
is represented by exactly one grid cell.

```{r no-merge-check}
count_check(no_merge, no_merge_grid)

no_merge_surface |>
  st_drop_geometry() |>
  select(cell, source_id, A, B)
```

# Case 2: Merge Required

Now create two tiny populated polygons along the left edge. With a four-by-two
grid, no grid-cell centroid falls inside either tiny polygon. The first-pass
rasterization therefore misses them.

The fallback step finds a representative interior point for each missed polygon
and identifies the grid cell containing that point. Because those cells are
already represented by source polygon 3, the tiny polygons are merged into
source polygon 3. The merged counts are then distributed over all grid cells
assigned to source polygon 3.

```{r merge-data}
merge_case <- st_sf(
  A = c(10, 0, 20, 0),
  B = c(0, 6, 4, 30),
  geometry = st_sfc(
    make_rect(0, 0, 0.2, 1),
    make_rect(0, 1, 0.2, 2),
    make_rect(0.2, 0, 2, 2),
    make_rect(2, 0, 4, 2)
  )
) |>
  source_to_sf()

merge_grid <- surface_from_polygons(merge_case, nrow = 2, ncol = 4)
merge_surface <- surface_to_sf(merge_grid)
```

```{r merge-source-map}
plot_sources(merge_case, "Source polygons: merge required")
```

```{r merge-grid-map}
plot_surface(merge_surface, "Grid surface after fallback merges")
```

The fallback metadata reports which source polygons were missed by the initial
centroid assignment and which fallback cells they used. In this example, source
polygons 1 and 2 are merged into the represented source polygon occupying those
fallback cells.

```{r merge-fallback}
tibble(
  missing_source_id = merge_grid$fallback$ids,
  fallback_cell = merge_grid$fallback$cells
)
```

The overall group counts are still preserved.

```{r merge-count-check}
count_check(merge_case, merge_grid)
```

But the source-to-grid bookkeeping changes. Source polygons 1 and 2 are not
represented as their own grid zones. Their counts are included in source zone 3,
and then spread over the four cells assigned to source zone 3.

```{r merge-zone-check}
source_totals <- merge_case |>
  st_drop_geometry() |>
  select(source_id, A, B)

surface_totals <- merge_surface |>
  st_drop_geometry() |>
  group_by(source_id) |>
  summarise(A = sum(A), B = sum(B), .groups = "drop")

source_totals
surface_totals
```

# Pycnophylactic Smoothing Uses the Merged Grid

For `surface = "pycno"`, the same grid construction happens first. The
pycnophylactic smoother then operates on the merged grid and preserves the
counts of the represented grid zones. This keeps the smoother coherent: it does
not receive unrepresented source polygons after the fact.

```{r pycno-merge}
merge_pycno <- surface_from_polygons(
  merge_case,
  nrow = 2,
  ncol = 4,
  surface = "pycno"
)
merge_pycno_surface <- surface_to_sf(merge_pycno)
```

```{r pycno-map}
plot_surface(merge_pycno_surface, "Pycnophylactic surface after fallback merges")
```

The total counts remain unchanged after smoothing.

```{r pycno-count-check}
count_check(merge_case, merge_pycno)
```

# Using the Surface in spseg()

In ordinary use, call `spseg()` directly. The same surface construction is used
internally before local environments and segregation indices are calculated.

```{r spseg-surface-call}
grid_result <- spseg(
  merge_case[, c("A", "B")],
  surface = "grid",
  nrow = 2,
  ncol = 4,
  bands = c(1, 2),
  output = "full"
)

pycno_result <- spseg(
  merge_case[, c("A", "B")],
  surface = "pycno",
  nrow = 2,
  ncol = 4,
  bands = c(1, 2),
  output = "full",
  converge = 1
)

as.data.frame(grid_result, what = "indices")
as.data.frame(pycno_result, what = "indices")
```

